Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-05-10 papers/histogram/sad.tex
final edit of paper

\begin{abstract}
  \davidsays{Let's look at the abstract together.}
  We present a novel Monte-Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a dynamic update factor (SAD-$\gamma$ or SAD)
  which dynamically adjusts the update factor $\gamma$ during the course of
  the simulation. We test this MC method on the square-well fluid and
  compare the convergence time and average entropy error for a variety of weight-based
  Monte-Carlo methods. SAD rapidly converges to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.
\end{abstract}

@@

\subsection{$1/t$-Wang-Landau}
The error saturation of Wang-Landau algorithm algorithm can be
corrected by modifying the update factor such that it does not
decrease too quickly.  Belardinelli and Pereyra raised the question as
to whether the update factor should be decrease by $\nicefrac12$ or
some other factor for each update~\cite{belardinelli2007fast}. They
implemented a schedule that enforced that if the update factor
$\gamma$ is ever less than $\nicefrac{N_S}{t}$ then the update factor
is set to $\nicefrac{N_S}{t}$ where $N_S$ is the number of energy
states and the histogram is no longer tracked. Employing this
scheduler, they found that the error saturation is avoided since the
correct density of states is approached asymptotically as
$t^{-\frac12}$~\cite{belardinelli2008analysis}. Zhou et. al further
confirmed that the WL algorithm never converges exponentially and
succesfully bounded the statistical error between $t^{-\frac12}$ and
$1/t$~\cite{zhou2008optimal}.

Schneider et. al outlines minor refinements to the $1/t$-WL algorithm
including $\nicefrac{N_S}{t}$ scaling and switching from standard WL
to $1/t$-WL when the update factor $\gamma <
\nicefrac{N_S}{t}$~\cite{schneider2017convergence}. \davidsays{Above
  we say that 1/t-WL does this.  One must be wrong?} As per the
original $1/t$ implementation~\cite{belardinelli2007fast}, the update
factor is decreased once all $N_S$ energy states are visited at least
once (i.e. $H(E) \neq 0$) effectively avoiding the concept of
`flatness'.

@@

The methods SAD, WL, and $1/t$-WL compensate for the smaller
displacement distance by reducing $\gamma$ more slowly, as can be seen from
Figure~\ref{fig:fast-slow-gamma}b and ~\ref{fig:fast-slow-gamma}d.
The update factors take approximately 10x longer to reach steady-state
for the slower translation scale. Because of this update behavior,
these methods are less sensitive to the choice of displacement
distance than SAMC is.  \davidsays{This paragraph could benefit from
  going through again.}

@@

\davidsays{I got to here.}

@@

\section{Conclusions}
Both SAD and 1/t-WL converge more rapidly than Wl or SAMC. SAD requires
the user to specify a temperature range rather than an energy range as 1/t-WL
does. Both WL and 1/t-WL require a prior simulation in order to determine the energy range
of interest. WL saturates in convergence as discussed in the introduction and
requires the user to optimize the choice of parameters to avoid this such as
increasing the flatness criterion.  SAMC converges for a reasonable choice of
$t_0$ but this parameter can be difficult to tune especially across different
energy bins, displacement distances, or systems. Additionaly, SAMC does not
converge as rapidly as either SAD or 1/t-WL even for the best choice of $t_0$.

We have introduced a new algorithm SAD (Stochastic Approximation with a
Dynamic $\gamma$) that effectively samples the entire energy space for
a variety of system sizes.  Also, SAD proves effective in comparison
with other Monte-Carlo methods at converging to the correct density of
states.  The ease of implementation and the absence of user-defined
parameters makes SAD an ideal algorithm. SAD converges as quickly as
SAMC (in the case of an ideally chosen $t_0$ for our simulations. SAD
doesn't suffer from the weakness of SAMC, i.e. the user choosing
parameters that could lead to the simulation not converging in a
reasonable amount of moves.

SAD decreases $\gamma$ in a controlled way that prevents the algorithm
from taking too long to converge (as in the case of SAMC with a poor choice
of $t_0$) or failing to converge such as is the case with WL for complex
systems. A major advantage is that the user does not need to determine a
priori how quickly the update factor $\gamma$ should decrease as the
method SAD dynamically determines this.

We note that SAD should be tested on additional systems, and the
convergence of the method should be examined.  Also, a rigorous
mathematical justification for our choice of the update factor is
necessary to ensure the most optimal convergence of the algorithm.

\section{Acknowledgement}

>>>>

\begin{abstract}
  We present a new Monte-Carlo algorithm based on the Stochastic
  Approximation Monte Carlo (SAMC) algorithm for directly calculating
  the density of states. The proposed method is Stochastic
  Approximation with a Dynamic update factor (SAD)
  which dynamically adjusts the update factor $\gamma$ during the course of
  the simulation. We test this method on the square-well fluid and
  compare the convergence time and average entropy error for several
  related
  Monte-Carlo methods. We find that both the SAD and $1/t$-Wang-Landau ($1/t$-WL)
  methods rapidly converge to the
  correct density of states without the need for the user to specify an
  arbitrary tunable parameter $t_0$ as in the case of SAMC.  SAD requires
  as input the temperature range of interst, in contrast to
  $1/t$-WL, which requires that the user identify the interesting range
  of energies.  Thus SAD is more convenient when the range of energies is
  not known in advance.
\end{abstract}

@@

\subsection{$1/t$-Wang-Landau}
The error saturation of Wang-Landau algorithm algorithm can be
corrected by modifying the update factor such that it does not
decrease too quickly.  Belardinelli and Pereyra raised the question as
to whether the update factor should be decrease by $\nicefrac12$ or
some other factor for each update~\cite{belardinelli2007fast}. They
implemented a schedule that enforced that if the update factor
$\gamma$ is ever less than $\nicefrac{1}{t}$ then the update factor
is set to $\nicefrac{1}{t}$ and the histogram is no longer tracked. Employing this
scheduler, they found that the error saturation is avoided since the
correct density of states is approached asymptotically as
$t^{-\frac12}$~\cite{belardinelli2008analysis}. Zhou et. al further
confirmed that the WL algorithm never converges exponentially and
succesfully bounded the statistical error between $t^{-\frac12}$ and
$1/t$~\cite{zhou2008optimal}.

Schneider et. al outlines minor refinements to the $1/t$-WL algorithm
including $\nicefrac{N_S}{t}$ scaling and switching from standard WL
to $1/t$-WL when the update factor $\gamma <
\nicefrac{N_S}{t}$~\cite{schneider2017convergence}. As per the
original $1/t$ implementation~\cite{belardinelli2007fast}, the update
factor is decreased once all $N_S$ energy states are visited at least
once (i.e. $H(E) \neq 0$) effectively avoiding the concept of
`flatness'.  In this work, we refer to this refined algorithm as $1/t$-WL,
with the update factor asymptoting to $\nicefrac{N_S}{t}$.

@@

The methods SAD, WL, and $1/t$-WL compensate for the smaller
displacement distance by reducing $\gamma$ more slowly, as can be seen from
Figure~\ref{fig:fast-slow-gamma}b and ~\ref{fig:fast-slow-gamma}d.
The update factors take approximately 10$\times$ longer to reach steady-state
for the smaller displacement distance. Because of this update behavior,
these methods are less sensitive to the choice of displacement
distance than SAMC is.

@@

@@

\section{Conclusions}
We have introduced a new algorithm that effectively samples the
energy space corresponding to a desired range of temperatures for a
variety of system sizes. 
%
We find that both SAD and $1/t$-WL demonstrate excellent and robust
convergence.
They both converge more rapidly than SAMC, and unlike WL consistently
converge to the correct density of states. SAD requires
the user to specify a temperature range of interest rather than an
energy range of interest as $1/t$-WL does.  For use cases in which
a range of desired temperatures is known, this will make the SAD method
considerably more convenient.

We find that SAMC converges for a reasonable choice of
$t_0$ but this parameter can be difficult to tune especially across
significantly differing systems.  We find that even simple
changes to the Monte Carlo moves can have a dramatic effect on the
range of practical $t_0$ values.  Additionaly, SAMC does not
converge as rapidly as either SAD or $1/t$-WL even for the best choice of $t_0$,
when a relatively small range of energies is required, because it always
simulates all possible energies.

%We note that SAD should be tested on additional systems, and the
%convergence of the method should be examined.  Also, a rigorous
%mathematical justification for our choice of the update factor is
%necessary to ensure the most optimal convergence of the algorithm.

%\section{Acknowledgement}
